--- 
title: "Data I/O"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document: default
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.path = '../assets/')
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

We can write a dataset to disk in comma-separated format using the `write.csv` function. 
Using our previous "states" example...

```{r eval=FALSE}
write.csv(states, file="states.csv")
```

Note the named argument to indicate the filename. We could have specified any
other folder/directory by passing a path. In the previous call, the file will be
written to our current working directory. We can see which one it is with:

```{r}
getwd()
```

and we can use the `setwd()` to set it. 

Unsurprisingly, we can read our dataset back using `read.csv`.
```{r eval=FALSE}
states <- read.csv("states.csv")
states
```

What about other delimiters and even formats? For other delimiters, we can use
the more general function `read.table` and `write.table` that allows us to
specify which delimiter we want to use. Actually, `read.csv` is just
`read.table` with a predefined delimiter

```{r eval=FALSE}
states <- read.table("states.csv", sep=",")
```

and we could have, for instance, defined that our input is tab separated by specifying 

```{r eval=FALSE}
states <- read.table("states.csv", sep="\t")
```

The most common format for R uses the extension `.RData` (or `RDS`), using the functions
`save` (`saveRDS`) and `load` (`readRDS`):

```{r eval=FALSE}
save(states, file="states.RData")
load("states.RData")
states
```

But R can also read (and sometimes write) data in other binary formats: data
coming from Stata, SAS, SPSS, or even Excel. The functions to handle these
foreign formats are provided in the `foreign` package that comes with `R` but a
much better alternative is the `haven` package which gives us access to
functions like `read_sas` to read SAS sas7bdat files or `read_stata` to read
Stata dta files.

This package is not provided with R but we need instead to install it and
loaded, which gives us a good opportunity to look at importing new functionality
into R.

We first need to install the library using:

```{r}
install.packages("haven")
```

The function will hit a CRAN mirror, download the file for the package that we
want, and perform the installation routine (which includes a number of checks).
Now the package is available in our system, we can load it into our session:

```{r}
library(haven)
```

For instance, we can now load an _uncompressed_ SAS file using 

```{r}
path <- system.file("examples", "iris.sas7bdat", package="haven")
read_sas(path)
```

or we could read a Stata file using 

```{r}
path <- system.file("examples", "iris.dta", package="haven")
read_stata(path)
```

As a matter of fact, one of the good things about `R` is that it interacts
nicely with many other programs. For instance, it can read Excel spreadsheets in
several different ways. One option is `readxl` library, which
does not require Java. So first, like before, we need to install the package and
load it:

```{r}
install.packages("readxl")
library(readxl)
```

Now we will be able to call a function like `read_excel`

```{r include=FALSE}
path <- readxl_example("geometry.xls")
```

```{r}
read_excel(path)
```

We can also query data from databases. There are a number of packages for this,
but the one I prefer is `RODBC` (some database systems may be incompatible and 
require a different package).

```{r eval=FALSE}
library(RODBC)

db_str <- sprintf("driver={SQL Server};server=%s;database=%s;uid=%s;pwd=%s",
                   server, database, username, password)                   
db_handle <- odbcDriverConnect(db_str)

data <- sqlQuery(dbhandle, query_str)

odbcClose(db_handle)
```

Each of these function has a number of options (Do we want to use a catalog file
for SAS? In which sheet is the data in our Excel file? How to deal with user
generated missing values in SPSS?) and of course we can interact with many other
formats (`feather` for data exchange with Python, fixed width files, hd5, ...),
streams (for real-time analysis), ... but this is sufficient for now.
